---
title: "Matrix Variate Normal Distributions"
author: "Geoffrey Thompson"
date: "February 3, 2018"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Matrix Normal Distributions}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
bibliography: matrixnormal.bib  
---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment="")
library(matrixdist)
```

## Matrix Variate Distributions

Suppose there is a two-dimensional grid of observations: each observation is randomly distributed somehow, perhaps each spot on the grid has its own mean, there's a covariance relation across the rows that doesn't vary across the columns and a covariance relation across the columns that doesn't vary across the rows. One case where this might come up is multivariate time series: at each point in time we are dealing with a multivariate observation (perhaps normally distributed or not) and then there is a covariance structure across time to account for. We may be interested in drawing from such random variables, calculating densities, or fitting parameters for such distributions based on observations of those distributions. Here I have some functions for doing so in the case of matrix variate normal distributions. For more details about matrix variate distributions in general and matrix variate normal distributions in particular, see @gupta1999matrix, whose results I rely on for much of this presentation.

## Matrix Variate Normal Distributions

A random matrix $\mathbf{X}$ $(p \times n)$ with a matrix variate normal distribution is parameterized by a mean matrix $\mathbf{M}$ $(p \times n)$ and covariance matrices $\mathbf{U}$ $(p \times p)$ (determining covariance among rows) and $\mathbf{V}$ $(n \times n)$ (determining covariance among columns) and has the following probability density function (@Matrix_normal_distribution):
\[p(\mathbf{X}\mid\mathbf{M}, \mathbf{U}, \mathbf{V}) = \frac{\exp\left( -\frac{1}{2} \, \mathrm{tr}\left[ \mathbf{V}^{-1} (\mathbf{X} - \mathbf{M})^{T} \mathbf{U}^{-1} (\mathbf{X} - \mathbf{M}) \right] \right)}{(2\pi)^{np/2} |\mathbf{V}|^{n/2} |\mathbf{U}|^{p/2}} \]
Here is a useful fact for drawing observations from a matrix variate normal distribution: suppose $\mathbf{X}$ is distributed as a matrix variate normal distribution with mean matrix $\mathbf{0}$ and covariance matrices $\mathbf{I_n}$ and $\mathbf{I_p}$. Then for a mean matrix $\mathbf{M}$ and linear transformations $\mathbf{L}$ and $\mathbf{R}$ of appropriate dimensions, if 
\[Y = \mathbf{M} + \mathbf{L}\mathbf{X}\mathbf{R}\]
then $\mathbf{Y}$ is matrix variate normally distributed with parameters $\mathbf{M}, \mathbf{LL}^T, \mathbf{R}^T\mathbf{R}$. 

Matrix variate random variables can then be generated by specifying $\mathbf{M}$ and $\mathbf{L}$ and $\mathbf{R}$ or by specifying $\mathbf{U}$ and $\mathbf{V}$. If the covariance matrices are provided, then Cholesky decomposition is performed to create the $\mathbf{L}$ and $\mathbf{R}$ matrices. If $\mathbf{L}$ and $\mathbf{R}$ matrices are provided, they are used to create $\mathbf{U}$ and $\mathbf{V}$ matrices which are then decomposed. If these are not specifed, the variances are presumed to be identity matrices by default and the means are presumed to be zero.

```{r mnorm}
set.seed(20180203)
x <- matrix(rnorm(6),nrow=3)
x
set.seed(20180203)
y <- rmatrixnorm(n = 1, mean=matrix(rep(0,6),nrow=3))
y
U <- 5 * diag(3) + 1
V <- matrix(c(2,0,0,.1),nrow=2)
mu = matrix(1:6,nrow=3)
set.seed(20180203)
z <- rmatrixnorm(n = 1, mean=mu,U=U,V=V)
z
mu + t(chol(U)) %*% y %*% chol(V)

```

When $n=1$, by default this returns a matrix. For $n>1$, by default, it will return a three-dimensional array indexed by the first coordinate. 

```{r noptions}
set.seed(20180203)
x <- rmatrixnorm(n = 1, mean=matrix(rep(0,6),nrow=3))
x
set.seed(20180203)
y <- rmatrixnorm(n = 100, mean=matrix(rep(0,6),nrow=3),list = TRUE)
y[[1]]
set.seed(20180203)
z <- rmatrixnorm(n = 100, mean=matrix(rep(0,6),nrow=3),array = TRUE)
z[1, , ]
class(x)
class(y)
class(z)
```

Densities are computed in much the same way as for other distributions. The defaults for the mean parameters are 0 and identity for the variance matrices. The `log` parameter sets whether to return the density on the log scale - by default this is `FALSE`.

```{r density}
set.seed(20180202)
A = rmatrixnorm(n=1,mean=matrix(c(100,0,-100,0,25,-1000),nrow=2),
L=matrix(c(2,1,0,.1),nrow=2))
dmatrixnorm(A,mean=matrix(c(100,0,-100,0,25,-1000),nrow=2),
  L=matrix(c(2,1,0,.1),nrow=2),log=TRUE )

```

## Parameter estimation for matrix variate normal distributions

Maximum likelihood estimation is possible for the matrix variate normal distribution (see @dutilleul1999mle or @2013arXiv1309.6609G for details). The pair of variance matrices are only identifiable up to a constant, so this function sets the first element of $\mathbf{U}$ to $1$.

There are two restrictions possible for the mean matrices: `row.mean` will force a common  mean within a row and `col.mean` will force a common mean within a column. Restrictions on $\mathbf{U}$ and $\mathbf{V}$ are possible with `row.variance` and `col.variance` commands. Currently the only option is to impose an AR(1) structure by providing the `AR(1)` option. This works by using `optim` to find the appropriate $\rho$ - it is not fast but if this is the true structure it will be better than an unstructured variance matrix. The $\rho$ parameter shouuld be $>0$ but is not forced to be non-negative.

In order to fit parameters for an $n \times p$ matrix, at least $\max (n/p,p/n) + 1$ observations are needed.

The function expects data in a format produced by the `rmatrixnorm` function.
The results are only the parameter estimates and do not include measures of uncertainty.

```{r mle,warning=FALSE}
set.seed(20180202)
A = rmatrixnorm(n=100,mean=matrix(c(100,0,-100,0,25,-1000),nrow=2),
   L=matrix(c(2,1,0,.1),nrow=2),list=TRUE)
results=mle.matrixnorm(A)
print(results)
results.fixrows = mle.matrixnorm(A, row.mean=TRUE)
print(results.fixrows)
# this failure is expected with misspecification! 

B = rmatrixnorm(n=50,mean=matrix(1:15,nrow=5),U=3*stats::toeplitz(.6^(1:5)))
mle.matrixnorm(B,row.variance="AR(1)",U=3*stats::toeplitz(.6^(1:5)))
mle.matrixnorm(B,U=3*stats::toeplitz(.6^(1:5)))

```

## References
